Motional Ground State Cooling Outside the Lamb-Dicke Regime

### 2017-07-25 paper.tex
More axis renaming

Our experiment begins by loading a single sodium atom into an optical tweezer from a MOT
for $250$ ms~\cite{Hutzler2017-LightShifts}, and has an overall repetition rate of $2.5$ Hz.
The tweezer is created by focusing a $700$ nm laser beam through an NA=$0.55$ objective to
create a focused beam with waist $0.7 \mu$m.
With $45$ mW of tweezer power, the trapping frequencies along the three axes are
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69(1), 430(4), 590(5)\}\ \text{\text{kHz}}$
where axis 1 is along the weakly confined axial direction and axis 2 and 3 are along the
tightly confined radial directions.
After attempting to load an atom from the MOT, an image is taken with $1.5$ ms exposure time to determine if it was a success.
At the same time, the atom is cooled via polarization gradient cooling (PGC),
which reduces the temperature of the single atom to $70$ $\mu$K,
corresponding to an initial mean motional state along the three different axes of
$\{\bar n_1, \bar n_2, \bar n_3\}=\{21(2),\, 3.4(2),\, 2.5(2)\}$, and therefore initial 3D ground state probability of $P_0=0.3$\%.
A beam with $\sigma^-$-polarization is used to initialize the atom into
the $|F=2, m_F=-2\rangle$ stretched state via optical pumping.
The hyperfine state detection is performed by pushing out atoms in the stretched state with a strong
beam on resonance with the $|F=2, m_F=-2\rangle$ to $|F'=3, m_{F'}=-3\rangle$ transition before
taking the second image to determine if the atom was pushed out.
The state preparation and detection have a combined fidelity of larger than $99$\%.

@@

For an atom in motional level $n$, the OP could result in motional-state redistribution
as shown in Fig.~\ref{f-ld}A. The state redistribution probability
is approximately proportional to the effective LD parameter
$\eta^{OP}_{eff}=\sqrt{2n+1}\eta^{OP}$, where $\eta^{OP}=k z_0=k \sqrt{\hbar/2m\omega}$,
$k$ is the wavenumber of the optical pumping beam, $z_0$ is the zero-point wavefunction spread,
$m$ is the mass of the atom, and $\hbar$ is the reduced Planck constant.
Such heating presents a major challenge for efficient cooling of single Na atoms
which feature a light mass and short D-line wavelengths.
For our trap frequencies and beam geometry,
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.602(5), 0.241(2), 0.206(1)\}$.
Combined with a relatively large PGC temperature gives initial
$\{\eta^{OP}_{1\eff},\eta^{OP}_{2\eff},\eta^{OP}_{3\eff}\} = \{4.0(1), 0.67(2), 0.50(1)\}$.
For the weak axial direction, $(\eta^{OP}_{1\eff})^2>1$ is far outside of the LD regime.
As a result, the average change of motional state per OP step is large,
which creates a large uncertainty on the motional state at the end of a full RSC step, possibly leading to runaway heating.
For example, the average change in motional states over the initial distribution
is $2.4$ in the axial direction, which is significantly greater than $0.89-0.95$
in previous experiments of RSC of Rb and Cs~\cite{Li2012,Kaufman2012,Thompson2013,Liu2017}.

Fortunately, the large LD parameters also provide opportunities to overcome OP heating.
The Raman transitions in our configuration have LD parameters of
$\{\eta^R_{1},\eta^R_{2},\eta^R_{3}\} = \{0.40(1), 0.341(2), 0.291(1)\}$,
which allow strong coupling to higher order sidebands for atoms in higher motional states
as calculated in Figure~\ref{f-ld}B.
The LD parameter for Raman transitions is defined as $\eta^R\equiv\Delta k z_0$ similar to OP
where $\Delta k$ being the wave vector difference
between the two beams that drives the Raman transition.
To offset heating from OP initially, higher-order Raman cooling sidebands can be utilized
to remove more motional quanta in a single cooling pulse.
Since the coupling strengths of different orders do not reach minima for the same pulse duration,
using multiple orders of cooling sidebands avoids accumulations of population
near the coupling minima.

@@

For the weaker axial direction, cooling is much more challenging
due to the large $\eta^{OP}_{1eff}\approx 4.0(2)$.
We observe up to 8th-order Raman cooling sidebands initially,
which indicates population in highly-excited motional states.
Nevertheless, our cooling sequence works efficiently as all the cooling sidebands are suppressed
after RSC (Fig.~\ref{f-axial}).
The ground state population calculated using the ratio of first-order sideband heights yields
$92(3)$\%. We also perform Rabi flopping on the carrier(Figure~\ref{f-axial}B)
and extracted a ground state population of $91(2)$\%, in agreement with the other method.
For the first-order heating sideband Rabi flopping (Figure~\ref{f-axial}C),
we observed additional decoherence that is more pronounced due to the slower Rabi frequency.
The decoherence time scale is consistent with magnetic field fluctuations of $1.5$ mG
that we measure independently in the lab, which produces a Zeeman shift of $\sim 3$ kHz.

@@

\fxnote{[The distinction between 77 and 66 percent is confusing, we should discuss it]}

>>>>

Our experiment begins by loading a single sodium atom into an optical tweezer from a MOT
for $250$ ms~\cite{Hutzler2017-LightShifts}, and has an overall repetition rate of $2.5$ Hz.
The tweezer is created by focusing a $700$ nm laser beam through an NA=$0.55$ objective to
create a focused beam with waist $0.7 \mu$m.
With $45$ mW of tweezer power, the trapping frequencies along the three axes are
$\{\omega_x,\omega_y,\omega_z\}/2\pi = \{430(4), 590(5), 69(1)\}\ \text{\text{kHz}}$
where axis z is along the weakly confined axial direction and axis x and y are along the
tightly confined radial directions.
After attempting to load an atom from the MOT, an image is taken with $1.5$ ms exposure time to determine if it was a success.
At the same time, the atom is cooled via polarization gradient cooling (PGC),
which reduces the temperature of the single atom to $70$ $\mu$K,
corresponding to an initial mean motional state along the three different axes of
$\{\bar n_x, \bar n_y, \bar n_z\}=\{3.4(2),\, 2.5(2),\, 21(2)\}$, and therefore initial 3D ground state probability of $P_0=0.3$\%.
A beam with $\sigma^-$-polarization is used to initialize the atom into
the $|F=2, m_F=-2\rangle$ stretched state via optical pumping.
The hyperfine state detection is performed by pushing out atoms in the stretched state with a strong
beam on resonance with the $|F=2, m_F=-2\rangle$ to $|F'=3, m_{F'}=-3\rangle$ transition before
taking the second image to determine if the atom was pushed out.
The state preparation and detection have a combined fidelity of larger than $99$\%.

@@

For an atom in motional level $n$, the OP could result in motional-state redistribution
as shown in Fig.~\ref{f-ld}A. The state redistribution probability
is approximately proportional to the effective LD parameter
$\eta^{\textrm{OP}}_{\textrm{eff}}=\sqrt{2n+1}\eta^{\textrm{OP}}$, where $\eta^{\textrm{OP}}=k z_0=k \sqrt{\hbar/2m\omega}$,
$k$ is the wavenumber of the optical pumping beam, $z_0$ is the zero-point wavefunction spread,
$m$ is the mass of the atom, and $\hbar$ is the reduced Planck constant.
Such heating presents a major challenge for efficient cooling of single Na atoms
which feature a light mass and short D-line wavelengths.
For our trap frequencies and beam geometry,
$\{\eta^{\textrm{OP}}_x,\eta^{\textrm{OP}}_y,\eta^{\textrm{OP}}_z\} = \{0.241(2), 0.206(1), 0.602(5)\}$.
Combined with a relatively large PGC temperature gives initial
$\{\eta^{\textrm{OP}}_{x,\textrm{eff}},\eta^{\textrm{OP}}_{y,\textrm{eff}},\eta^{\textrm{OP}}_{z,\textrm{eff}}\} = \{0.67(2), 0.50(1), 4.0(1)\}$.
For the weak axial direction, $(\eta^{\textrm{OP}}_{z,\textrm{eff}})^2>1$ is far outside of the LD regime.
As a result, the average change of motional state per OP step is large,
which creates a large uncertainty on the motional state at the end of a full RSC step, possibly leading to runaway heating.
For example, the average change in motional states over the initial distribution
is $2.4$ in the axial direction, which is significantly greater than $0.89-0.95$
in previous experiments of RSC of Rb and Cs~\cite{Li2012,Kaufman2012,Thompson2013,Liu2017}.

Fortunately, the large LD parameters also provide opportunities to overcome OP heating.
The Raman transitions in our configuration have LD parameters of
$\{\eta^R_x,\eta^R_y,\eta^R_z\} = \{0.341(2), 0.291(1), 0.40(1)\}$,
which allow strong coupling to higher order sidebands for atoms in higher motional states
as calculated in Figure~\ref{f-ld}B.
The LD parameter for Raman transitions is defined as $\eta^R\equiv\Delta k z_0$ similar to OP
where $\Delta k$ being the wave vector difference
between the two beams that drives the Raman transition.
To offset heating from OP initially, higher-order Raman cooling sidebands can be utilized
to remove more motional quanta in a single cooling pulse.
Since the coupling strengths of different orders do not reach minima for the same pulse duration,
using multiple orders of cooling sidebands avoids accumulations of population
near the coupling minima.

@@

For the weaker axial direction, cooling is much more challenging
due to the large $\eta^{\textrm{OP}}_{z,\textrm{eff}}\approx 4.0(2)$.
We observe up to 8th-order Raman cooling sidebands initially,
which indicates population in highly-excited motional states.
Nevertheless, our cooling sequence works efficiently as all the cooling sidebands are suppressed
after RSC (Fig.~\ref{f-axial}).
The ground state population calculated using the ratio of first-order sideband heights yields
$92(3)$\%. We also perform Rabi flopping on the carrier(Figure~\ref{f-axial}B)
and extracted a ground state population of $91(2)$\%, in agreement with the other method.
For the first-order heating sideband Rabi flopping (Figure~\ref{f-axial}C),
we observed additional decoherence that is more pronounced due to the slower Rabi frequency.
The decoherence time scale is consistent with magnetic field fluctuations of $1.5$ mG
that we measure independently in the lab, which produces a Zeeman shift of $\sim 3$ kHz.

@@

\fxnote{The distinction between 77 and 66 percent is confusing, we should discuss it}
